---
title: "Amplicon Read Depth"
author: "Kim Siegmund"
date: "10/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GenomicAlignments)
library(Rsamtools)
library(tidyverse)
library(writexl)
```

# Load the amplicon regions targeted for sequencing

```{r readBedfile}
load("data/targetgr.rda")
```

Filenames for this processing run.
```{r runnames}
#datadir=c("~kims/Google Drive File Stream/My Drive/methylation bam files 9-19")
datadir=c("~kims/Google Drive File Stream/My Drive/methylation ampliseq bam files 12-19")
writedatafn=c("~kims/GitHub/bsseq/data/readDepth-38-amplicons-dec2019-experimental-controls.xlsx")
```

Now let's count reads that cover this region using: countOverlaps(galp,targetgr).

# Darryl's BAMs:  

Grab filenames.

```{r bamfilenames}
bamfname=dir(datadir,pattern="*\\.bam$")
bamfiles=file.path(datadir,bamfname)
nbam <- length(bamfiles)
```

Quick summary:
```{r bamfilesummary}
# shorten names
sslist <- strsplit(bamfname,"Galaxy")
nm <- map_chr(sslist, ~.x[1])
nm <- substr(nm,6,stop = 100000L)

for (i in 1:length(nm)) {
  print(nm[i])
  quickBamFlagSummary(bamfiles[i], main.groups.only=TRUE)
}
```

Here are some data frames to store the counts matrix of reads covering target.
```{r makedfs}
cnts.df <- data.frame(matrix(NA,nrow=3,ncol=(1+length(nm))))
colnames(cnts.df) <- c("Amplicon",nm)
rownames(cnts.df) <- c("NReads","NOnTarget","pctOnTarget")

overlaptable.df <- data.frame(matrix(NA,nrow=38,ncol=(1+length(nm))))
colnames(overlaptable.df)=c("Amplicon",nm)
overlaptable.df$Amplicon <- 1:38
```

I'm going to count 1st reads, which all map to +-strand. I can revisit this decision later.

\it{From R documentation: Prepare the ScanBamParam object to perform the filtering, we’ll 
get rid of the PCR or optical duplicates (flag bit 0x400 in the SAM format,
see the SAM Spec 1 for the details), as well as reads not passing quality 
controls (flag bit 0x200 in the SAM format).} 
```{r scanbamflag}
 flag1 <- scanBamFlag(isFirstMateRead=TRUE, isSecondMateRead=FALSE,
 isDuplicate=FALSE, isNotPassingQualityControls=FALSE)
 param1 <- ScanBamParam(flag=flag1, what="seq")
```

Now we will read *.bam file and count reads covering amplicons. 
```{r readGAlign}
for (i in 1:length(nm)) {
  galr1 <- readGAlignments(bamfiles[i], use.names=TRUE, param=param1)
  r1ontarget <- findOverlaps(ranges(galr1),ranges(targetgr),
                           minoverlap=min(width(targetgr)))
  if (length(table(countOverlaps(ranges(galr1),ranges(targetgr),
                    minoverlap=min(width(targetgr))))) > 2)   print("WARNING: Read covers more than amplicon")

  lgr1 <- length(galr1)
  lr1on<- length(r1ontarget)
  cnts.df["NReads",nm[i]] <- lgr1
  cnts.df["NOnTarget",nm[i]] <- lr1on
  cnts.df["pctOnTarget",nm[i]] <- lr1on/lgr1

  tb <- table(subjectHits(r1ontarget))
  overlaptable.df[names(tb),nm[i]] <- tb
}
```

Combine 2 tables into a single file to export.
```{r tout}
readDepth <- rbind.DataFrame(cnts.df,overlaptable.df)
readDepth <- cbind.data.frame(Summary = rownames(readDepth),
                             readDepth)
#write_xlsx(readDepth,path ="data/readDepth-38-amplicons.xlsx")
write_xlsx(readDepth,path = writedatafn)
```

```{r sI}
sessionInfo()
```

