---
title: "Clip 38 regions to CpGs"
author: "Kim Siegmund"
date: "10/14/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GenomicAlignments)
library(tidyverse)
library(writexl)
library(BSgenome.Hsapiens.UCSC.hg19)
```

# Load annotation information

Load the 38 amplicons & CpG positions.
```{r loadannot}
load("data/cgpos.rda")
load("data/targetgr.rda")
load("data/targetseqs.rda")
```

# Clip the ranges to 1st and last CpG position
Create and save new CpG positions to correspond to position in clipped target regions

```{r cliprange}
cliptargetgr <- targetgr
firstcg <- map_int(cgpos,pluck,1)
lastcg <- map_int(cgpos,~pluck(.x, length(.x)))
start(cliptargetgr) <- start(targetgr) - 1 + firstcg
end(cliptargetgr) <- start(targetgr) - 1 + lastcg
cliptargetgr[c(1:2,38)]
```

Let's check these against the human ref genome. I'll add 1 base to see if the targets with a C in the final position is a CpG.
```{r viewclippedtargets}
#to view complete CpGs, I will include 1 base 3' of amplicon.
clipgrplus1 <- cliptargetgr
end(clipgrplus1) <- end(cliptargetgr) + 1
vi = suppressWarnings(Views(Hsapiens,clipgrplus1))
clipseqplus1 <- as(vi,"DNAStringSet")
clipseqplus1
```

```{r cgposclippedgr}
firstcg <- map_int(cgpos,pluck,1)  
cgposafterclip <- map2(cgpos,firstcg, ~ .x -.y + 1)
cgposafterclip[1:3]
```

Save these results for calling DNA methylation haplotypes. Save the trailing G position because the minus strand-first mate in pair reads will need to be reverse engineered.
```{r saveme}
save(clipgrplus1,file="data/clipgrplus1.rda")
save(cgposafterclip,file="data/cgposafterclip.rda")
```

```{r sI}
sessionInfo()
```

