---
title: "Summarize Data By ReadDepth"
author: "Ks"
date: '`r paste("First created on 10/30/2019. Updated on", Sys.Date())`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Create Scatter Diagrams of Diversity measure vs Read Depth

We're going to sample our reads to different depths, and measure diversity of the sampled reads. The two diversity measures we consider are (1) the number of unique haplotypes and (2) hamming distance. We then plot diversity measure vs read depth. We expect that the diversity will increase with read depth, reaching a plateau that captures the biological diversity in the sample.
```{r libraries, message=FALSE}
if(!require("readxl")) install.packages("readxl")
library(tidyverse)
library(readxl)
library(ggplot2)
```

Read all Excel files into a list.
```{r readxlsx}
#datadir=c("~kims/GitHub/newbsseq/data/cmethhaps/experimental-ctls")
datadir=c("~kims/GitHub/runbsseq/data/cmethhaps/20200620")
#datadir=c("~kims/GitHub/newbsseq/data/cmethhaps/exprmntl-ctls-filt")
#datadir=c("~kims/GitHub/newbsseq/data/cmethhaps/darrylsfafiles9-19biscuit")

xlsxfiles=dir(datadir,pattern="*\\.xlsx$")
xlsxfiles

load("~kims/GitHub/runbsseq/data/cgpos.rda")
```

Now I want to curate certain files for individual plots.
```{r groupfiles}
#nm <- substr(xlsxfiles,1,3)
#set1 <- paste0(1:9,"_M")
#set1idx <- which(is.element(nm,set1))
#xlsxfiles[set1idx]
```

Here I can look at smaller sets of samples.  Set2 subsets on dish 1, passages 18, 70, 105.
```{r gpsubsetfiles}
#sname <- sub(".xlsx","",xlsxfiles[set1idx])
#shname <- strsplit(sname,("[+]"))
#sname <- map_chr(shname, ~.x[3])
#getdishno <- strsplit(sname,("[_]"))
#posdishno <- which(map_chr(getdishno, ~.x[2])==1)
#sname <- factor(sname,levels=sname)[posdishno]

#set2idx <- set1idx[posdishno]
#xlsxfiles[set2idx]

#setidx <- set1idx
#setidx <- set2idx
#setidx <- c(2,3,1)
setidx <- c(1,2,3)
xlsxfiles[setidx]
```

```{r hdfunction}
avghamdist <- function(hapmatrix,nh){
     nr <- length(nh)
     # count number of hamdist of 0 for replicate haplotypes
     npairssame <- map_dbl(nh,~choose(.x,2))
     
     hamdist <- dist(t(hapmatrix),method="manhattan")
     # compute weights using replicates of each haplotype
     distwgt <- NA
     for (i in 1:(nr-1))  {
            distwgt <- c(distwgt,nh[i]*nh[-(1:i)])      
          }
     distwgt <- distwgt[-1]
     distwgt
     hd <- sum(hamdist*distwgt)/(sum(distwgt)+sum(npairssame))
     hd
}
```

Here's the function that will sample the reads at different read depths, and compute diversity measures on each subsample. 
```{r getdivmeasuresinsubsamples}
# x is a vector of haplotypes
# ssize is a vector of increasing readdepths
sumssample <- function(x,ssize){
      ns <- length(ssize)
      uhap      <- rep(NA,ns)
      avghd     <- rep(NA,ns)
      for (i in 1:ns) {
        # sample haplotypes  
        sample.x <- 
               x %>%
                dplyr::sample_n(ssize[i])
        # get number of unique haplotypes  
        uhap[i]  <- 
               sample.x %>%
                dplyr::n_distinct()  
             
        # get counts of each unique haplotype
        nhaps <- 
            sample.x %>%
              group_by(htype) %>%
              dplyr::summarise(nhap = length(htype),.groups="keep") 
        
        # compute hamming distance
        names(nhaps$htype)  <- paste0("v",c(1:length(nhaps$htype)))
       hapdf <- map_dfc(nhaps$htype,
                       ~as.numeric(unlist(strsplit(.x,split=""))))
       hapmat <- as.matrix(hapdf)
       avghd[i] <- avghamdist(hapmat,nhaps$nhap)
      }
      ncpgs <- nrow(hapmat)
      avghd <- avghd/ncpgs
      sumstats <- bind_cols(uhap=uhap,avghd=avghd)
}
```
 
Let's read in the data and summarize diversity for subsets of the data, looping over the amplicons.
```{r analyzedata}
#number of samples in this set
nxlsx <- length(setidx)
#sample name
sname <- sub(".xlsx","",xlsxfiles[setidx])
#shname <- strsplit(sname,("[+]"))
#sname <- map_chr(shname, ~.x[3])
sname <- factor(sname,levels=sname)

#amplicon names (for Excel sheets)
amplname <- paste0("amplicon",1:38)

# prototype amplicon 1, then 1-2, then loop over all 38
#an<-1
#for (an in 1:2) {  #test on shorter list

#amplicons without dropouts
qampl <- c(1:7,12:31,33:37)
#require > 4 CpGs for inclusion in haplotype diversity analysis
qampl <- qampl[map_int(cgpos,length)[qampl]>4]


#Loop over the amplicons, skipping dropouts
for (an in qampl) {
#for (an in 1:2) {

#Create a list, each element is the list of haplotypes for one sample  
  amplst <- vector(mode = "list", length = nxlsx)
  names(amplst) <- sname
for (i in 1:nxlsx) {
    amplst[[i]] <- read_excel(file.path(datadir,xlsxfiles[setidx[i]]),
                            sheet = amplname[an])
    amplst[[i]] <- amplst[[i]] %>%
                    dplyr::select(htype)
}

# Summarize Read depth for each sample
readdepth <- map_int(amplst,nrow)
print(amplname[an])
print("Total Read Depths")
names(readdepth) <- sname
print(readdepth)

#Since each sample has a different read depth, we'll compute a sample-specific vector of sampled read depths using percentiles with a step-size of 5% (i.e. 10%, 15%, ...,100%).  
depthlst <- vector(mode = "list", length = nxlsx)
rpct <- seq(0.1:1,by=0.05)
for (i in 1:nxlsx)   depthlst[[i]] <- floor(readdepth[i]*rpct)
#for (i  in  1:nxlsx) depthlst[[i]] <- c(10, 20, 30)
#print("(Quantiles of) Read Depths, in step-sizes of 5%")
#print(depthlst)

# Compute No. of unique haplotypes for randomly sampled subsets of size depthlst
set.seed(32)
slst <- map2_dfr(amplst,depthlst,~sumssample(.x,.y))

#Now let's plot it.   
sumdat <- bind_cols(uhap = slst$uhap,
                    avghd = slst$avghd,
                    depth = unlist(depthlst),
                    samplename = rep(sname,each=length(depthlst[[1]])))

par(mfrow=c(1,2))
p1 <- 
  ggplot(sumdat, aes(depth,uhap,colour = samplename)) +
  geom_line(aes(linetype = samplename)) +
  labs(title = paste0(amplname[an],",  ",length(cgpos[[an]])," CpGs"), 
       x = "read depth", y = "# unique haplotypes")
plot(p1)
p2 <- 
  ggplot(sumdat, aes(depth,avghd,colour = samplename)) +
  geom_line(aes(linetype = samplename)) +
  labs(title = paste0(amplname[an],",  ",length(cgpos[[an]])," CpGs"), 
       x = "read depth", y = "Avg Hamming Distance") +
    ylim(0,0.5)
plot(p2)
}
```

```{r sessionInfo}
sessionInfo()
```