---
title: "Summarize Data By ReadDepth"
author: "Ks"
date: "10/30/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Create a Scatter Diagram of Unique Haps vs Read Depth

We're going to sample our reads to different depths, and count the unique number of haplotypes we find for each depth. Then plot # unique haplotypes vs read depth.
```{r libraries, message=FALSE}
if(!require("readxl")) install.packages("readxl")
library(tidyverse)
library(readxl)
```

Read all Excel files into a list and count the number of reads per sample.
```{r readxlsx}
datadir=c("~kims/GitHub/bsseq/data/cmethhaps")
xlsxfiles=dir(datadir,pattern="*\\.xlsx$")

nxlsx <- length(xlsxfiles)
amplst <- vector(mode = "list", length = nxlsx)
for (i in 1:nxlsx) 
  amplst[[i]] <- read_excel(file.path(datadir,xlsxfiles[i]))

map_int(amplst,nrow)
```

Now let's subset the data to amplicon 1, get the total number of reads in amplicon 1 for each sample. 
```{r getamplicon1}
thisamp <- 1
amp1lst <- vector(mode = "list", length = nxlsx)

for (i in 1:nxlsx) 
  amp1lst[[i]] <- amplst[[i]] %>%
                    dplyr::filter(amplicon==thisamp) %>%
                    dplyr::select(htype)

readdepth <- map_int(amp1lst,nrow)
readdepth
```

Since each sample has a different read depth, we'll compute a sample-specific vector of sampled read depths using percentiles with a step-size of 5% (i.e. 5%, 10%, 15%, ...,100%).  
```{r sampledepths}
depthlst <- vector(mode = "list", length = nxlsx)
rpct <- seq(0.1:1,by=0.05)
for (i in 1:nxlsx)   depthlst[[i]] <- floor(readdepth[i]*rpct)
depthlst
```

Now we compute the number of unique haplotypes after we sample at random from our reads.
```{r subsample}
sumssample <- function(x,ssize){
      ns <- length(ssize)
      uhap      <- rep(NA,ns)
      for (i in 1:ns) {
            uhap[i] <-
               x %>%
                dplyr::sample_n(ssize[i]) %>%
                dplyr::n_distinct()        
      }
      uhap
}
  
set.seed(32)
slst <- map2(amp1lst,depthlst,~sumssample(.x,.y))
slst
```

Now let's plot it!

```{r matplot,echo=FALSE}
hmat <- matrix(unlist(slst),ncol=nxlsx)  
xmat <- matrix(unlist(depthlst),ncol=nxlsx)
matplot(xmat,hmat,
        main=paste("Amplicon:",thisamp),
        xlab="read depth",ylab="# unique haplotypes")
```

```{r sessionInfo}
sessionInfo()
```